---
title: "Cleaning and wrangling"
author: "O'Hara"
format: html
---

## Setup

```{r}
#| warning: false
#| message: false

library(dplyr)
library(tidyr)
library(readr)
```

## Read data

```{r}
#| warning: false
#| message: false

catch_original <- read_csv('https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1')
```

## Explore data

```{r}
colnames(catch_original)

head(catch_original)

summary(catch_original)

unique(catch_original$Region)
```


## Clean and Reshape

### Use `select` function to choose columns

```{r}
catch_data <- catch_original |>
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

colnames(catch_data)
```

### Quality check

```{r}
glimpse(catch_data)
```


### Replace values in a column using `mutate`

```{r}
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))
```

Finding rows with NA values

```{r}
i <- which(is.na(catch_clean$Chinook))

catch_data[i, ] ### or filter(catch_clean, is.na(Chinook))
```

Fixing the I in Chinook column

```{r}
catch_clean <- catch_data %>%
  mutate(Chinook = if_else(condition = (Chinook == 'I'), 
                           true = '1', 
                           false = Chinook),
         Chinook = as.numeric(Chinook))
```

### Reshape data with `pivot_longer` and `pivot_wider`

```{r}
catch_long <- catch_clean %>%
  pivot_longer(cols = -c(Region, Year),
               names_to = 'species',
               values_to = 'catch')

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
  pivot_wider(names_from = species,
              values_from = catch)
```


### Rename columns using `rename`

```{r}
catch_long <- rename(catch_long, catch_thousand = catch)
```

### Add a new column using `mutate`

```{r}
catch_long <- catch_long %>%
  mutate(catch = catch_thousand * 1000) %>%
  select(-catch_thousand)

head(catch_long)
```


### Demonstrate split-apply-combine workflow

```{r}
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarize(mean_catch = mean(catch))

head(mean_region)

n_region <- catch_long %>%
  group_by(Region) %>%
  summarize(n = n())

head(n_region)
```


### Use `filter` to subset by rows
```{r}
sse_catch <- catch_long %>%
  filter(Region == 'SSE')
```


### Sort data frame using `arrange`

### Split and combine values in columns using `separate` and `unite`


```{r}
mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = if_else(Chinook == 'I', '1', Chinook),
         Chinook = as.numeric(Chinook)) %>%
  pivot_longer(-c(Region, Year), names_to = 'species', values_to = 'catch') %>%
  mutate(catch = catch * 1000) %>%
  group_by(Region) %>%
  summarize(mean_catch = mean(catch)) %>%
  arrange(desc(mean_catch))
```

